---
title: "temp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{temp}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Packages

```{r setup}
library(sparrowtelomeres)
library(readxl)
library(reshape2)
```



## Raw data

The raw data is available via the publisher's website (https://royalsocietypublishing.org/doi/suppl/10.1098/rsbl.2015.0559) in a table in a word document called "rsbl20150559supp1.docx".  This file is saved in the folder /inst/extdata.

Data tables can be extracted directly from Word documents using
docxtractr::docx_extract_all_tbls()

(This code would set up the file name; however, I didn't follow through this this)
```{r, eval = F}
file.name <- "rsbl20150559supp1.docx"
file.full <- here::here("/inst/exdata",file.)
```


Data was copied and pasted from the Word document into an Excel spreadsheet into a sheet called "orig".

Set up the file name
```{r, eval = F}
file.name <- "Meillere_et_al_2015_telomere_data.xlsx"
file.full <- here::here("inst/extdata",file.name)
```

Read in the datasheet
```{r, eval = F}
dat.orig <- readxl::read_xlsx(file.full,sheet = "orig")
```

## Clean data

```{r, eval = F}
head(dat.orig)
summary(dat.orig)
summary(factor(dat.orig$sex))
```

## Set factors

```{r, eval = F}
dat.orig$nest <- as.factor(dat.orig$nest)
dat.orig$treatment <- as.factor(dat.orig$treatment)
dat.orig$sex <- as.factor(dat.orig$sex)
dat.orig$fledging <- as.factor(dat.orig$fledging)
```

## Rename object

Give it a useful name

```{r, eval = F}
telosfull <- dat.orig
```


Save as data .rda object
```{r, eval = F}
use_data(telosfull, overwrite = T)
```

Save as .csv

```{r, eval = F}
file.full <- here::here("/inst/extdata","telosfull.csv")
write.csv(telosfull, file.full, row.names = F)
```


## Fledging success data

The authors used a binomial glm using the number fledged per nest for fledging success models.

Note: in the data provided, fledging success appears to be recorded at the brood level, even though in methods they report modeling the propotion of nestling fledged, which implies nestling-level data.

```{r, eval = F}
#reshape data
telosfull[,c("nest","fledging")]
fledgedata <- reshape2::dcast(data = telosfull,
      formula = nest + treatment ~  fledging)

#rename columns
names(fledgedata)[3:4] <- c("fledged.no","fledged.yes")

#calculate total number of fledglings in brood
## see note above about nest-level success
fledgedata$brood.size <- fledgedata$fledged.no + fledgedata$fledged.yes

# code nest as successful or not
fledgedata$nest.success <- ifelse(fledgedata$fledged.yes > 0, 1, 0)

#create data object
use_data(fledgedata, overwrite = T)

#creat .csv
file.full <- here::here("/inst/extdata","fledgedata.csv")
write.csv(fledgedata,file.full,row.names = F)
```


## Subset data
```{r, eval = F}


dat.ext  <- readxl::read_xlsx(file.,sheet = "subset.ext2")
dat.sub  <- readxl::read_xlsx(file.,sheet = "subset.MF")

```

